---
title: "Reproducible Research Project I"
author: "Vishal"
date: "04/06/2020"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

```{r}
if(!file.exists('activity.csv')){
  unzip('activity.zip')
}

dfActivity <- read.csv('activity.csv')
```


## Histogram of the total number of steps taken each day

A new dataframe is created to count the number of steps on a particular date. And then the same is plotted using ggplot library.

```{r}
library(ggplot2)
library(dplyr)

dfActivity2 <- dfActivity %>%
  group_by(date) %>%
  summarise(totalSteps = sum(steps, na.rm = TRUE))

ggplot(dfActivity2, aes(totalSteps)) +
  geom_histogram(binwidth = 500) +
  ggtitle("Total number of steps taken each day")
```

## Mean and median number of steps taken each day

```{r}
stepsMean <- mean(dfActivity2$totalSteps)
stepsMedian <- median(dfActivity2$totalSteps)

cat("Average steps taken: ", stepsMean)
cat("median of steps taken: ", stepsMedian)
```

## Time series plot of the average number of steps taken
```{r}
dfActivity3 <- dfActivity %>%
  group_by(interval) %>%
  summarise(meanSteps = mean(steps, na.rm = TRUE))

ggplot(dfActivity3, aes(interval, meanSteps)) +
  geom_line(color = "blue", size = 0.75) +
  ggtitle("Mean number of steps taken in 5 minutes interval") +
  xlab("5 minute interval") +
  ylab("Mean number of steps")
```

## The 5-minute interval that, on average, contains the maximum number of steps

```{r}
maxAvgSteps <- which.max(dfActivity3$meanSteps)
maxAvgBy5MinInterval <- dfActivity3[maxAvgSteps, 'interval']
maxAvgBy5MinInterval <- gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", maxAvgBy5MinInterval)
cat("The 5-minute interval that, on average, contains the maximum number of steps: ", maxAvgBy5MinInterval)
```

## strategy for imputing missing data

The folowing startegy is used for imputing the missing data:

1. The original datafrme is copied in another dataframe.

2. A function is written which fills the NA fields with the mean of the present data.

3. Then the column in the new dataframe is updated with the imputed data.

```{r}
dfActivityImputed <- dfActivity
imputeFunction <- function(x) {
  replace(x, is.na(x), mean(x, na.rm = TRUE))
}

dfActivityImputed$steps <- imputeFunction(dfActivity$steps)

head(dfActivityImputed)
```

## Histogram of the total number of steps taken each day after missing values are imputed

```{r}
dfActivityImputedDailySteps <- dfActivityImputed %>%
  group_by(date) %>%
  summarise(totalSteps = sum(steps))

ggplot(dfActivityImputedDailySteps, aes(totalSteps)) +
  geom_histogram(binwidth = 500) +
  ggtitle("Number of steps taken per day (Imputed data)") +
  xlab("Total steps")
```

## Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}
dfActivityImputed$date <- as.Date(dfActivityImputed$date, '%Y-%m-%d')

weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
dfActivityImputed$day <- factor((weekdays(dfActivityImputed$date) %in% weekdays1), 
                   levels=c(FALSE, TRUE), labels=c('weekend', 'weekday')) 

dfWeekends <- dfActivityImputed %>%
  group_by(interval) %>%
  filter(day == 'weekend') %>%
  mutate(meanSteps = mean(steps, na.rm = TRUE))

dfWeekdays <- dfActivityImputed %>%
  group_by(interval) %>%
  filter(day == 'weekday') %>%
  mutate(meanSteps = mean(steps, na.rm = TRUE))                   

df_new <- rbind(dfWeekdays, dfWeekends)

ggplot(df_new, aes(interval, meanSteps)) +
  geom_line() +
  facet_grid(day~.) +
  ggtitle("Comparison of avg. number of steps taken on weekends and weekdays") +
  ylab("Average steps")
```